### 2023-12-03 thesis/MSECE_thesis.tex
Thesis edits.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the availability of large
natural language corpora corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain,
the aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the
most prevalent of the traditional models are Hidden Markov models, Gaussian Mixture models, and/or in conjunction with word lattices, N-grams, and/or
rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpora in the general domain is either
WikiText or the BookCorpus which contain trillions of tokens whereas, for aviation, the largest corpus is Air Traffic Control Complete with
approximately 26 hours of labeled data amounting to just over 300,000 tokens\footnote{Air Traffic Control Complete has approximately 70 hours of
    audio data, however, about 26 hours of the 70 are labeled.}. This disparity in data availability is exacerbated by the fact that aviation English
uses a specialized vocabulary in addition to specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T)
medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to transcribe/label data as it requires transcribers to have domain
knowledge and experience in order to create effective transcriptions. The medium of communication along with the environment in which the
communication occurs i.e.~a noisy and possibly low fidelity medium\footnote{Depending on the equipment used to transmit and receive R/T
    communications.} combined with a noisy and active communication environment---an active flight deck (influenced by wind, engine noise, etc.~) or
an active Air Traffic Control (ATC) tower (other people/controllers conversing or communicating in the background)---and a rapid rate of speech
\cite{paltridge_handbook_2013}, further contributing to the difficulty in understanding and therefore transcribing aviation English.

% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the SQuAD benchmark has
recorded results consistently better than human since 2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that
these NLP algorithms could have high utility in the aviation domain. If they can achieve better-than-human performance in the aviation domain, and be
used in conjunction with pilot and air traffic controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate
errors during normal operations.

The methods and results laid out in this work were done with the intention of creating a streamlined process for training and testing deep learning
based NLP methods, specifically automatic speech recognition and text-based language models. The results from the language models suggest that the
volume of text used to train the language models is not yet sufficient to produce results akin to those in the conversational and/or literary English
domains. However, the automatic speech recognition models were able to achieve near state-of-the-art results and, according to the literature review,
better than other deep learning-based approaches in the aviation domain.

\section{Notation and Terminology}
% TODO: proper writing, bulleted list for now
This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
below.

Set notation is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts. With that being said, the
shorthand below is defined
\begin{equation*}
    A; x := A \cup \{x\}
\end{equation*}

\noindent
as adding the element $x$, of its own set, to the set $A$.

The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
\begin{itemize}
    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (section
          \ref{sec:tokenizers})
    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
\end{itemize}

\section{Literature Review}\label{sec:lit_review}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
state-of-the-art (28.4 BLEU on WMT 2014 English-to-German; 41.8 BLEU on WMT English-to-French datasets)\cite{vaswani_attention_2017}. Transformer
architectures have been found to be extremely effective at learning representations and understandings of languages to predict token probabilities as
opposed to transforming one language into another\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at
other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks,
respectively)\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~proposed pretraining methods such as masked language
modeling and next sentence prediction to improve natural language understanding for downstream tasks \cite{devlin_bert_2019}, which was further
expanded upon by Liu et al.~by training longer, with larger batch sizes, and removing the next sentence prediction task \cite{liu_roberta_2019}.

% tokenizer algorithms
In language modeling and natural language processing tasks more generally, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by
the introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding have been used to solve OOV problems while simultaneously minimizing
vocabulary sizes and maximizing the likelihood of tokens in sequences \cite{wu_googles_2016,schuster_japanese_2012,sennrich_neural_2016}.

@@

% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification,
etc. A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language
modeling is used to transcribe speech
\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
and word lattices have also been used in the aviation domain alongside semi-supervised approaches to automatic speech recognition to leverage
unlabeled data, however, none of the developed methodologies use neural language models for their applications
\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.

% differences in language (transcripts vs literary english), narrow model selection
The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then
training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text
domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models
\cite{raffel_exploring_2020}. Although, Yadlowsky et al.~recently showed that if problem domains are sufficiently different, model pretraining on data
considered out-of-domain will have negative effects on model performance \cite{yadlowsky_pretraining_2023}.

\section{Data Sources}\label{sec:data_source}
There are four corpora used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
transcriptions, audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in
sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in section \ref{sec:data_analysis}.

\subsection{Air Traffic Control Complete}\label{sec:atcc}
The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of VHF
antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types of frequencies
monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora~\cite{godfrey_air_1994}.
% TODO: data length/size

\subsection{ATCO2}\label{sec:atco2}
The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
transcriptions. The speech recordings and transcripts are crowd sourced from volunteers. The dataset was created and labeled, in part, as an English
language detection corpus, however, the labeling also includes transcripts of speech segments which makes it practical for ASR and language model
development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
% TODO: data length/size

\subsection{Air Traffic Control Simulation}\label{sec:atcosim}
The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
The data was recorded at a air traffic control simulation center in Bretigny-sur-Orge, France, specifically the EUROCONTROL Experimental Centre.
In this data, only the controllers' voices are included and therefore the transcripts only include the controller's side of each interaction.
The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was
transcribed, it was reviewed and corrected and any remaining problems were reviewed by an operational air traffic controller and
resolved\cite{hofbauer_atcosim_2008}.
% TODO: data length/size

\subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
unified\cite{smidl_air_2019}.
% TODO: data length/size

\section{Data Processing}\label{sec:data_processing}
All of the corpora listed above (see~\ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal
being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).

@@

% 22 spelling errors
Manually reviewing every sample in the corpus is unrealistic considering that there are over 50,000 data samples and would be subject the same
human error that introduced those errors in the first place. Instead, the frequency of occurrence of tokens in the corpora combined with manual review
thereafter, was used to detect and correct the most common errors. Tokens with the lowest frequency across corpora were manually reviewed and
corrected where necessary. An \(A \rightarrow B\) mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
corrected or replaced token. A total of 22 unique errors were detected and subsequently corrected or removed using this method, although multiple
instances of those errors typically occurred.

@@

\begin{table}[h!]
    \centering
    \begin{tabular}{l p{0.65\linewidth}}
        \toprule
        Phenomenon               & Definition                                                                                                              \\
        \midrule
        Hesitations              & Instances of hesitation for which the specific realization is not defined (usually some special token is used instead). \\
        \midrule
        Partial \& cut-off words & Partially pronounced words are indicated by a hyphen character where the pronunciation of the word is missing.          \\
        \midrule
        Filled pauses            & Hesitations with a verbal component that has been described or otherwise defined by the corpus/annotator.               \\
        \bottomrule
    \end{tabular}
    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
    \label{tab:phenomena_definitions}
\end{table}

\begin{table}[h!]
    \centering
    \begin{tabular}{l l}
        \toprule
        Phenomenon                   & Convention for representation \\
        \midrule
        Hesitations                  & [HES]                         \\
        \midrule
        Partial \& cut-off words     & -                             \\
        \midrule
        Filled pauses (open-mouth)   & uh, um, ah                    \\
        \midrule
        Filled pauses (closed-mouth) & hm, mhm                       \\
        \bottomrule
    \end{tabular}
    \caption{Conventions for transcript representation/translation of phenomena to keep transcripts consistent across corpora.}
    \label{tab:phenomena_conventions}
\end{table}

@@

\begin{table}[h!]
    \centering
    \begin{tabular}{l r r r r}
        \toprule
        \textbf{Corpus} & \textbf{Samples} & \textbf{Mean Sequence Length (\(\pm \sigma\))} & \textbf{Unique Tokens} & \textbf{Tokens} \\
        \midrule
        ATCC            & 29,862           & \(11.90 \pm 7.22\)                             & 2,209                  & 355,473         \\
        ATCO2           & 874              & \(12.28 \pm 4.12\)                             & 786                    & 10,733          \\
        ATCOSIM         & 9,544            & \(11.23 \pm 4.12\)                             & 823                    & 107,153         \\
        ZCU CZ ATC      & 14,435           & \(10.05 \pm 5.50\)                             & 2,983                  & 145,107         \\
        \midrule
        Total           & 54,715           & \(11.30 \pm 6.36\)                             & 5,040                  & 618,466         \\
        \bottomrule
    \end{tabular}
    \caption{The number of samples, mean sequence length, number of unique tokens, and total number of tokens by corpus, including the cumulative
        total across all corpora.}
    \label{tab:corpora_stats}
\end{table}

@@

\subsection{Outlier Detection and Analysis}\label{sec:outliers}
The Local Outlier Factor (LOF) algorithm~\cite{breunig_lof_2000} was used to detect and remove outliers in the corpora for further analysis. The
sequence lengths of the data samples were used to determine outliers. Since all four corpora are aggregated, LOF analysis was performed on the
aggregated data instead of on the individual corpora.

Sci-Kit Learn's implementation of the LOF
algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the K-Nearest Neighbors algorithm and Minkowski distance for distance
metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were
detected as outliers and removed for analysis and are summarized in table~\ref{tab:outlier_stats}.

\begin{table}[h!]
    \centering
    \begin{tabular}{l | r}
        \toprule
        \multicolumn{2}{c}{Outlier Properties}   \\
        \midrule
        \textbf{Mean Sequence Length}    & 47.30 \\
        \textbf{Std of Sequence Length}  & 11.82 \\
        \textbf{Minimum Sequence Length} & 38    \\
        \textbf{Maximum Sequence Length} & 73    \\
        \textbf{\# Samples}              & 47    \\
        \bottomrule
    \end{tabular}
    \caption{Mean, standard deviation, minimum, and maximum sequence length of samples detected as outliers.}
    \label{tab:outlier_stats}
\end{table}

@@

The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
RoBERTa pretraining and downstream task training thereafter, it was initially decided to eliminate the next sentence prediction (NSP) task as it was
used in BERT, however, their results suggest that the success of the MLM-only approach was due to the increase in batch and overall data
size~\cite{liu_roberta_2019}. Taking this into account, along with the very limited amount of data available in the domain of aviation an NSP
pretraining task was implemented where possible\footnote{Only about half of the data (only the data from ATCC, since it is continuous recordings of
    the airspace) is in a format that enables us to extract the turn taking phenomenon that is present in most air-traffic communications.} to improve
its pretraining results. MLM and NSP pretraining are explained more in-depth in section \ref{sec:bert}, below.

@@

\begin{equation}\label{eq:ngram_prob_wj}
    P(w) = \frac{\textit{occurrences}(w)}{D_{tokens}}
\end{equation}

\noindent
where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\textit{occurrences}(w)$ is the number of times of a specific
token appears in the corpus, and $D_{tokens}$ is total number of tokens in the corpus, $D$. Computationally, the process of determining the
probabilities of the individual tokens in the corpus is typically referred to as training or fitting the N-gram model.

@@

\textbf{Next Sentence Prediction}. NSP is meant to support downstream tasks, while MLM conditions the model to represent tokens within sentences, NSP
is meant to condition the model to understand relationships between sentences/sequences. Since BERTs training set consists primarily of literary works
and excerpts from Wikipedia, the model simply predicts whether two sentences appear together (one after the other) in the original text. Instead of a
single input sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token, these
are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.

For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
and NSP objectives, input sequences always start with a classification token and end with a separation token (see figure \ref{fig:bert_mlm_input_example}).

\begin{figure}[h!]
    \includegraphics[width=\linewidth]{figures/BERT_MLM_input.png}
    \caption{An example of an input sequence for a masked language modeling pretraining task. Modified from Devlin et al.~\cite{devlin_bert_2019}.}
    \label{fig:bert_mlm_input_example}
\end{figure}

>>>>

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the availability of large
natural language corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain, the
aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the most
prevalent of the traditional models are N-grams, word lattices, and/or rule-based approaches depending on the application and operating environment of
the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the
aviation domain. By comparison, the largest corpora in the general domain is either WikiText or the BookCorpus which contain trillions of tokens
whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to
transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective transcriptions. This, in
addition to the following facets, further contribute to the difficulty of transcription: (1) the medium of communication, (2) the environment in which
the communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive
    communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4)
an active Air Traffic Control (ATC) tower and/or channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.

% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question
Answering Dataset (SQuAD) benchmark has recorded results consistently better than human since
2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
aviation domain. If they can achieve better-than-human performance in the aviation domain, and be used in conjunction with pilot and air traffic
controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate errors during normal operations.

The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep learning
based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the language
models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the results show
that language models trained ``from scratch'' on in-domain data perform better than model trained on out-of-domain data and transfer learned on
in-domain data i.e.~pretrained BERT and RoBERTa models transfer learned to aviation English.

\section{Notation and Terminology}
This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
below.

Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field
    of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms,
to convey mathematical concepts. With that being said, the shorthand below is defined
\begin{equation*}
    A; x := A \cup \{x\}
\end{equation*}

\noindent
as the union of the set $A$ and a set containing the element $x$. The equation below can be read as adding the element $x$ to the set $A$:
\begin{equation*}
    A \leftarrow A; x
\end{equation*}

The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
\begin{itemize}
    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (Section
          \ref{sec:tokenizers})
    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
\end{itemize}

\section{Literature Review}\label{sec:lit_review}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
state-of-the-art \cite{vaswani_attention_2017}. Transformer architectures have been found to be extremely effective at learning representations and
understandings of languages to predict token probabilities as opposed to transforming one language into another
\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP-related tasks such as prompt completion
and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019}proposed pretraining methods such as masked language modeling and
next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.

% tokenizer algorithms
In language modeling and natural language processing tasks, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by the
introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding (BPE) have been used to solve OOV problems while simultaneously minimizing
vocabulary sizes and maximizing the likelihood of tokens in sequences \cite{wu_googles_2016,schuster_japanese_2012,sennrich_neural_2016}.

@@

% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification,
etc. A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language
modeling is used to transcribe speech
\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
and word lattices have also been used in the aviation domain alongside semi-supervised approaches to automatic speech recognition to leverage
unlabeled data. However, none of the developed methodologies use neural language models for their applications
\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.

% differences in language (transcripts vs literary english), narrow model selection
The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then
training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text
domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models
\cite{raffel_exploring_2020}. Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different,
model pretraining on data considered out-of-domain will have negative effects on model performance.

\section{Data Sources}\label{sec:data_source}
There are four corpora used in this thesis. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
transcriptions, audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in
sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in Section
\ref{sec:data_analysis}.

\subsection{Air Traffic Control Complete}\label{sec:atcc}
The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of Very
High Frequency (VHF) antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types
of frequencies monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora
\cite{godfrey_air_1994}.
% TODO: data length/size; samples, hours, tokens

\subsection{ATCO2}\label{sec:atco2}
The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
transcriptions. The speech recordings and transcripts are crowd sourced from volunteers. The dataset was created and labeled, in part, as an English
language detection corpus, however, the labeling also includes transcripts of speech segments which makes it practical for ASR and language model
development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
% TODO: data length/size; samples, hours, tokens

\subsection{Air Traffic Control Simulation}\label{sec:atcosim}
The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
The data was recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France. In this corpus, only the
controllers' voices are included and therefore the transcripts only include the controller's side of each interaction. The audio data was transcribed
by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was transcribed, it was reviewed and
corrected and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}.
% TODO: data length/size; samples, hours, tokens

\subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
unified\cite{smidl_air_2019}.
% TODO: data length/size; samples, hours, tokens

\section{Data Processing}\label{sec:data_processing}
All of the corpora listed above (see Section \ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal
being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).

@@

% 22 spelling errors
Manually reviewing every sample in the corpus is unrealistic considering that there are over 50,000 data samples and would be subject the same
human error that introduced those errors in the first place. Instead, the frequency of occurrence of tokens in the corpora combined with manual review
thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation
(\ref{eq:token_frequency_occurrence}).

\begin{equation}\label{eq:token_frequency_occurrence}
    f_w = \frac{\phi_w(w)}{D_{tokens}}
\end{equation}

\noindent
Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and
corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed, and if an erroneous token, $A$, is found, it is replaced
with the corrected version, $B$. A total of 22 unique errors were detected and subsequently corrected or removed using this method, although multiple
instances of those errors typically occurred.

@@

\begin{table}[!t]
    \centering
    \begin{tabular}{l p{0.65\linewidth}}
        \toprule
        Phenomenon               & Definition                                                                                                              \\
        \midrule
        Hesitations              & Instances of hesitation for which the specific realization is not defined (usually some special token is used instead). \\
        \midrule
        Partial \& cut-off words & Partially pronounced words are indicated by a hyphen character where the pronunciation of the word is missing.          \\
        \midrule
        Filled pauses            & Hesitations with a verbal component that has been described or otherwise defined by the corpus/annotator.               \\
        \bottomrule
    \end{tabular}
    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
    \label{tab:phenomena_definitions}
\end{table}

\begin{table}[!t]
    \centering
    \begin{tabular}{l l}
        \toprule
        Phenomenon                   & Convention for representation \\
        \midrule
        Hesitations                  & [HES]                         \\
        \midrule
        Partial \& cut-off words     & -                             \\
        \midrule
        Filled pauses (open-mouth)   & uh, um, ah                    \\
        \midrule
        Filled pauses (closed-mouth) & hm, mhm                       \\
        \bottomrule
    \end{tabular}
    \caption{Conventions for transcript representation/translation of phenomena to keep transcripts consistent across corpora.}
    \label{tab:phenomena_conventions}
\end{table}

@@

\begin{table}[!t]
    \centering
    \begin{tabular}{l r r r r}
        \toprule
        \textbf{Corpus} & \textbf{Samples} & \textbf{Mean Sequence Length (\(\pm \sigma\))} & \textbf{Unique Tokens} & \textbf{Tokens} \\
        \midrule
        ATCC            & 29,862           & \(11.90 \pm 7.22\)                             & 2,209                  & 355,473         \\
        ATCO2           & 874              & \(12.28 \pm 4.12\)                             & 786                    & 10,733          \\
        ATCOSIM         & 9,544            & \(11.23 \pm 4.12\)                             & 823                    & 107,153         \\
        ZCU CZ ATC      & 14,435           & \(10.05 \pm 5.50\)                             & 2,983                  & 145,107         \\
        \midrule
        Total           & 54,715           & \(11.30 \pm 6.36\)                             & 5,040                  & 618,466         \\
        \bottomrule
    \end{tabular}
    \caption{The number of samples, mean sequence length, number of unique tokens, and total number of tokens by corpus, including the cumulative
        total across all corpora.}
    \label{tab:corpora_stats}
\end{table}

@@

\subsection{Outlier Detection and Analysis}\label{sec:outliers}
The Local Outlier Factor (LOF) algorithm \cite{breunig_lof_2000} was used to detect and analyze outliers in the corpora. The sequence lengths of the
data samples were used to determine outliers. Since all four corpora are aggregated, LOF analysis was performed on the aggregated data instead of on
the individual corpora.

Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the
K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively
used for distance calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in table
\ref{tab:outlier_stats}.

\begin{table}[!t]
    \centering
    \begin{tabular}{l | r}
        \toprule
        \multicolumn{2}{c}{Outlier Properties}   \\
        \midrule
        \textbf{Mean Sequence Length}    & 47.30 \\
        \textbf{Std of Sequence Length}  & 11.82 \\
        \textbf{Minimum Sequence Length} & 38    \\
        \textbf{Maximum Sequence Length} & 73    \\
        \textbf{\# Samples}              & 47    \\
        \bottomrule
    \end{tabular}
    \caption{Mean, standard deviation, minimum, and maximum sequence length of samples detected as outliers.}
    \label{tab:outlier_stats}
\end{table}

@@

The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
RoBERTa pretraining and downstream task training thereafter, it was initially decided to eliminate the next sentence prediction (NSP) task as it was
used in BERT, however, their results suggest that the success of the MLM-only approach was due to the increase in batch and overall data
size~\cite{liu_roberta_2019}. Taking this into account, along with the very limited amount of data available in the domain of aviation an NSP
pretraining task was implemented where possible\footnote{Only about half of the data (only the data from ATCC, since it is continuous recordings of
    the airspace) is in a format that enables us to extract the turn taking phenomenon that is present in most air-traffic communications.} to improve
its pretraining results. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert}, below.

@@

\begin{equation}\label{eq:ngram_prob_wj}
    P(w) = \frac{\phi_w(w)}{D_{tokens}}
\end{equation}

\noindent
where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\phi_w(w)$ is the number of times of a specific
token appears in the corpus, and $D_{tokens}$ is total number of tokens in the corpus, $D$. Computationally, the process of determining the
probabilities of the individual tokens in the corpus is typically referred to as training or fitting the N-gram model.

@@

\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sentences, NSP is meant to condition the model to
understand relationships between sentences/sequences to further support downstream tasks. Since BERTs training set consists primarily of literary
works and excerpts from Wikipedia, the  model simply predicts whether two sentences appear together (one after the other) in the original text.
Instead of a single input sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification''
token, these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.

For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
and NSP objectives, input sequences always start with a classification token and end with a separation token (see figure
\ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a
separation token terminating each sequence (see figure \ref{fig:bert_nsp_input_example}).

\begin{figure}[!t]
    \includegraphics[width=\linewidth]{figures/BERT_MLM_input.png}
    \caption{An example of an input sequence for a masked language modeling pretraining task. Modified from Devlin et al.~\cite{devlin_bert_2019}.}
    \label{fig:bert_mlm_input_example}
\end{figure}
